---
title: "IDS section 404 post live session 11"
author: "Brian Kruse"
date: "11/20/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R markdown loads and uses the fpp package
```{/users/briankruse/desktop/live session 11 ids.R}
```


## Including Plots
Questions a) After plotting the time series we do not see
any seasonal fluctuations and we see an upward trend since 1980.

```{r v, echo=FALSE}
library(fpp) 
data(ukcars)
v <- ukcars
plot(v)
```

Use a classical decomposition to calculate the trend-cycle and seasonal indices. 


c)	Do the results support the graphical interpretation from part (a)? 
yes it does because we start with a downward tred and go up on the first two parts. The third one is a perfectly seasonally adjusted trend. The fourth one identifies random walk 
```{r fitd, echo=FALSE}

fitd <- decompose(v) 
plot(fitd)

```


d)	Compute and plot the seasonally adjusted data. 
```{r eeadj, echo=FALSE}
eeadj <- seasadj(fitd)
plot(eeadj)

```

e)	Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier? The outlier creates a spike in the plot.

```{r v2, echo=FALSE}
library(tseries)
v2 <- ts(c(v[1:54],
                   v[55]+500,v[56:113] ),
                 start=c(1977,1),frequency=4)
plot(v2)

```


f)	Does it make any difference if the outlier is near the end rather than in the middle of the time series? 
yes it makes a difference during decomposition
the trend flattens out on first display and last display

```{r fitdv, echo=FALSE}
fitdv <- decompose(v2) 
plot(fitdv)
```


g)	Use STL to decompose the series. 
```{r v2.stl, echo=FALSE}
v2.stl = stl(v2, s.window="periodic")

plot (v2.stl)
```
